Matrix Transposition and Symmetry: a case study on Superscalar Architectures\\

### 2024-11-27 latex/paper.tex
latex: pseudocode

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{blindtext}  % generate random text
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

@@

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{blindtext}  % generate random text
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[noend]{algpseudocode}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\iffalse % pseudocode
  \begin{algorithm}
    \caption{My Algorithm}\label{euclid}
    \begin{algoritmic}[1]
      \procedure{MyProcedure}{}
      % use \state and \textit{} for variables
      % use \text for text
      % use \gets for assignments
      % use \If \EndIf \Return \Else
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}
\endif

@@

In this report we benchmark different implementations of matTranspose
and checkSymm, comparing implicit and explicit parallelism
techniques. In particular, the benchmarks aim to increase cache and SIMD performance,
speedup with multiple threads and the impact of compile flags optimizations. Data will be compared through tables
and visually via graphs such as the roofline model. All benchmarks
will be tested against every matrix sixe from $2^2$ to $2^{12}$, we will assume the size is a multiple of 4 for ease of implementation.
We will now explain the methodology for matTranspose.
The algorithms being benchmarked take a $N \times N$ input square matrix, an output matrix with the same shape and the size of the matrix.
The following will be benchmarked:
\begin{itemize}
\item \textit{matTranspose}: naive implementation of matTranspose, iterating over values of the input matrix and storing them in the output matrix with rows and columns indexes swapped.

  \begin{algorithm}
    \caption{matTranspose}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTranspose}{$Matrix input, Matrix output, int size$}
        \For{\textit{r} \textbf{ in } \textit{0..size}}
            \State \For{\textit{c} \textbf{ in } \textit{0..size}}
            \State $output[c][r] = input[r][c]$
            \EndFor
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

    \item \textit{matTransposeColumns}: similar to \textit{matTranspose}, but iterating on the input matrix first by columns and then by rows.
  The purpose of this implementation is to test a different access pattern.

  \begin{algorithm}
    \caption{matTranspose columns}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTransposeColumns}{$Matrix input, Matrix output, int size$}
        \For{\textit{r} \textbf{ in } \textit{0..size}}
            \State \For{\textit{c} \textbf{ in } \textit{0..size}}
            \State $output[r][c] = input[c][r]$
            \EndFor
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \item{matTransposeHalf}: similar to \textit{matTranspose}, but the algorithm
  iterates only on the upper half of the input matrix. For each
  pair $(i,j)$ of indices for rows and columns in the upper half of the input matrix $M$, the output matrix O will
  store store $O[i][j]=M[j][i]$ and $O[j][i] = M[i][j]$.

  \begin{algorithm}
    \caption{matTranspose half}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTransposeHalf}{$Matrix input, Matrix output, int size$}
        \For{\textit{r} \textbf{ in } \textit{0..size}}
            \State \For{\textit{c} \textbf{ in } \textit{r..size}}
            \State $output[r][c] = input[c][r]$
            \State $output[c][r] = input[r][c]$
            \EndFor
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \item{matTransposeCyclic}: implementation of \cite{b13}, storing the matrix as a contiguous one-dimensional array of size $N*N$.

  \begin{algorithm}
    \caption{matTranspose cyclic}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTransposeCyclic}{$Matrix input, Matrix output, int size$}
        \For{\textit{n} \textbf{ in } \textit{0..(size*size)}}
        \State int \textit{i} = $\frac{n}{size}$
        \State int \textit{j} = $n \mod N$
        $T[n] = M[size * j + i]$
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \textit{matTransposeIntrinsic}: based on \textit{matrix\_transpose4x4} from the linux kernel \cite{b16}, revised to work for out-of-place float matrices on x86 architectures and implemented using intrinsics. \textit{transpose\_4x4\_f32\_intrinsic} transposes a 4 by 4 matrix using vector registers of 128 bits by doing the following steps: load 4 rows of 4 floats from the the input matrix, combine pairs of rows in an alternating pattern (interleaving), shuffle to get the transpose and store them in the output matrix. \textit{matTransposeIntrinsic} iterates over 4x4 blockw in the input matrix and saves the computed transposed block in the output matrix with rows and columns swapped.

  \begin{algorithm}
    \caption{matTranspose intrinsic}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{transpose\_4x4\_f32\_intrinsic}{float* src0, float* src1, float* src2, float* src3, float* dst0, float* dst1, float *dst2, float* dst3}
    \textbf{comment} Load rows  
    \_\_m128 row0 = \_mm\_loadu\_ps(src0);
    \_\_m128 row1 = \_mm\_loadu\_ps(src1);
    \_\_m128 row2 = \_mm\_loadu\_ps(src2);
    \_\_m128 row3 = \_mm\_loadu\_ps(src3);

    \textbf{comment} unpack and interleave rows
    \_\_m128 t0 = \_mm\_unpacklo\_ps(row0, row1);
    \_\_m128 t1 = \_mm\_unpackhi\_ps(row0, row1);
    \_\_m128 t2 = \_mm\_unpacklo\_ps(row2, row3);
    \_\_m128 t3 = \_mm\_unpackhi\_ps(row2, row3);

    \textbf{comment} Shuffle to form transposed rows
    \_\_m128 d0 = \_mm\_movelh\_ps(t0, t2);
    \_\_m128 d1 = \_mm\_movehl\_ps(t2, t0);
    \_\_m128 d2 = \_mm\_movelh\_ps(t1, t3);
    \_\_m128 d3 = \_mm\_movehl\_ps(t3, t1);

    \textbf{comment} Store transposed rows
    \_mm\_storeu\_ps(dst0, d0);
    \_mm\_storeu\_ps(dst1, d1);
    \_mm\_storeu\_ps(dst2, d2);
    \_mm\_storeu\_ps(dst3, d3);

      \EndProcedure
      \procedure{matTransposeIntrinsic}{$Matrix input, Matrix output, int size$}
        \For{int i = 0, i < size, i += 4}
        \For{int j = 0, j < size, j += 4}
        \State transpose\_4x4\_f32\_intrinsic(input[i][j],
        input[i+1][j], input[i+2][j], input[i+3][j], output[j][i], output[j+1][i], output[j+2][i], output[j+3][i])
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \textit{matTransposeIntrinsicCyclic}: same as \textit{matTransposeIntrinsic} but the matrix is saved as a one dimensional array similarly to \textit{matTransposeCyclic}
%\item matTranspose vectorization:
%\item matTranspose half vectorization:
\item \textit{matTransposeUnrollingOuter}: similar to \textit{matTranspose} but the outer loop is manually unrolled

  \begin{algorithm}
    \caption{matTransposeUnrollingOuter}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTransposeUnrollingOuter}{$Matrix input, Matrix output, int size$}
        \For{$int c = 0, i < size, i += 4$}
            \State \For{\textit{r} \textbf{ in } \textit{0..size}}
            \State $output[c][r] = input[r][c]$
            \State $output[c+1][r] = input[r][c+1]$
            \State $output[c+2][r] = input[r][c+2]$
            \State $output[c+3][r] = input[r][c+3]$
            \EndFor
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \textit{matTransposeUnrollingInner}: same as \textit{matTransposeUnrollingOuter} but the inner for loop is manually unrolled

\item \textit{matTransposeHalfUnrollingOuter}: same as \textit{matTransposeHalf} but the outer for loop is manually unrolled

\item \textit{matTransposeHalfUnrollingInner}: same as \textit{matTransposeHalf} but the inner for loop is manually unrolled

\item \textit{matTransposeCyclicUnrolled}: same as \textit{matTransposeCyclic} but the for loop in manually unrolled

  \begin{algorithm}
    \caption{matTranspose cyclic unrolled}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{matTransposeCyclicUnrolled}{$Matrix input, Matrix output, int size$}
        \For{\textit{n} \textbf{ in } \textit{0..(size*size)}}
        $T[n] = M[size * ((n+1) \mod size ) + (\frac{n}{size})]$
        $T[n+1] = M[size * ((n+1+1) \mod size ) + (\frac{n+1}{size})]$
        $T[n+2] = M[size * ((n+2+1) \mod size ) + (\frac{n+2}{size})]$
        $T[n+3] = M[size * ((n+3+1) \mod size ) + (\frac{n+3}{size})]$
        \EndFor
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \textit{matTransposeOmpX}: this set of algorithms are all derived from \textit{matTranspose} with the outer for loops parallelized using omp with the number of threads \textit{X} being a power of 2 from $2^1$ to $2^6$.

\item \textit{matTransposeOmpXCollapse}: similar to \textit{matTransposeOmpX} but the for loops are parellelized using \textit{collapse(2)}

\item \textit{matTransposeOmp16SchedX}: similar to \textit{matTransposeOmpX} with 16 threads using different schedule algorithms, namely: static, synamic, guided.
\end{itemize}

We will now explain the methodology for \textit{checkSymm}. The algorithms take an $N \times N$ square matrix and Its size are return true if the matrix is symmetrix, false otherwise.
The algorithms being benchmarked are the wollowing:
\being{itemize}
\item \textit{checkSymm}: naive implementation of checkSymm. Iterates over the upper half of the matrix and testing if every value is equal to the symmetric value.

  \begin{algorithm}
    \caption{checkSymm}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{checkSymm}{$Matrix input, int size$}
        \State $isSymm = true$
        \For{\textit{i} \textbf{ in } \textit{0..size}}
            \For{\textit{j} \textbf{ in } \textit{i..size}}
            \If $input[i][j] != input[j][i]$
            \State $isSymm = false$
        \EndFor
        \Return $isSymm$
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

% \item \textit{checkSymColumns}:

  % \item \textit{checkSymVectorization}
\item \textit{checkSymmUnrolledOuter}: similar to \textit{checkSymm} but the outer loop is manually unrolled.

  \begin{algorithm}
    \caption{checkSymmUnrolledOuter}\label{matTranspose}
    \begin{algoritmic}[1]
      \procedure{checkSymmUnrolledOuter}{$Matrix input, int size$}
        \State $isSymm = true$
        \For{\textit{i} \textbf{ in } \textit{0..size}}
            \For{\textit{j} \textbf{ in } \textit{i..size}}
            \If $input[i][j] != input[j][i]$
            \State $isSymm = false$
            \If $input[i+1][j] != input[j][i+1]$
            \State $isSymm = false$
            \If $input[i+2][j] != input[j][i+2]$
            \State $isSymm = false$
            \If $input[i+3][j] != input[j][i+3]$
            \State $isSymm = false$
        \EndFor
        \Return $isSymm$
      \EndProcedure
    \end{algoritmic}
  \end{algorithm}

\item \textit{checkSymmUnrollingInner}: similar to \textit{checkSymmUnrolledOuter} but the inner loop is manually unrolled instead of the outer loop.
\item \textit{checkSymmUnrollingOuterNoBranch}:
\item \textit{checkSymmUnrollingInnerNoBranch}:
\item \textit{checkSymOmpX}:
\item \textit{checkSymOmpXCollapse}:
\item \textit{checkSymm16SchedX}:
\end{itemize}
